%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% FRI NLP Final Report
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[fleqn,moreauthors,10pt]{ds_report}
\usepackage[english]{babel}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{array}
\graphicspath{{fig/}}

\JournalInfo{FRI Natural Language Processing Course 2026}
\Archive{Final project report}

\PaperTitle{
\begin{center}
Final Report\\[0.3em]
Domain-Specific Fitness NLP Assistant Using Retrieval-Augmented Generation
\end{center}
}
\Authors{Domen Pahole, Luka Rizman, Nina Trivi\'c}

\affiliation{\textit{Advisors: Ale\v{s} \v{Z}agar}}

\Keywords{Natural Language Processing, Retrieval-Augmented Generation, Fitness Assistant, Semantic Retrieval, Large Language Models}
\newcommand{\keywordname}{Keywords}

\Abstract{
We present a domain-specific fitness assistant built around Retrieval-Augmented Generation (RAG). The system combines a sentence-transformer retriever, a Qdrant vector database, and the Llama 3.2 1B model served through Ollama. Our main goal is not to maximize raw language fluency, but to improve grounding to a fixed exercise dataset and reduce unsupported recommendations. On an 11-query retrieval benchmark, the system reaches Hit@1 = 0.636, Hit@3 = 0.909, Hit@5 = 0.909, Precision@5 = 0.582, and MRR = 0.742. Qualitative comparison against the same 1B model without retrieval shows that RAG produces more dataset-aligned answers when retrieval succeeds, but also inherits retrieval failures. The main take-away is that, for this project, retrieval quality is the dominant bottleneck and the most promising direction for future improvement.
}

\begin{document}
\flushbottom
\maketitle
\thispagestyle{empty}

\section*{1 Introduction}

General-purpose LLMs are often fluent, but they are not reliable domain experts when answers depend on a specific local knowledge base. Our project addresses this problem in the fitness domain. We build a conversational assistant that answers questions about exercises, equipment, muscle groups, and training suggestions using a structured exercise dataset rather than relying only on the model's parametric knowledge.

The project follows the course topic of domain-specific AI assistants. We define a narrow use case, prepare a domain knowledge base, implement retrieval and generation components, and evaluate both retrieval quality and end-to-end answer behavior. The final system is designed to be easy to rerun locally or on HPC, with all dependencies and scripts available in the repository.

\section*{2 Data and System}

\textbf{Dataset.} We use the MegaGym exercise dataset, which contains 2,918 exercise entries. Each entry includes a title, description, exercise type, target body part, equipment, difficulty level, and optional rating fields. The dataset is public and linked in the repository instructions, so it is not duplicated in version control.

\textbf{Knowledge representation.} Each exercise is converted into a textual record containing the title, description, type, body part, equipment, difficulty level, and rating metadata. These records are embedded and stored in Qdrant.

\textbf{Retriever.} We use the sentence-transformer model \texttt{sentence-transformers/all-MiniLM-L6-v2}. Similarity search is performed in Qdrant with cosine distance. The default retrieval depth is top-$k=5$.

\textbf{Generator.} For answer generation we use \texttt{llama3.2:1b} through Ollama. This choice was motivated by the course constraint to use a small model that can also be deployed on HPC. The same model is used in both the baseline and RAG settings, so the comparison isolates the effect of retrieval rather than model size.

\textbf{Prompting.} In the RAG setting, the prompt instructs the model to use only retrieved exercise entries, prefer matches by body part, equipment, and difficulty, remove duplicates, and explicitly state when the dataset is insufficient. The baseline prompt receives only the user question.

\textbf{Implementation details.} The repository contains a Streamlit interface, indexing script, retrieval evaluation script, baseline-vs-RAG comparison script, and a unified evaluation runner. The system can run locally with Docker-based Qdrant or on Arnes HPC with local Qdrant storage and Ollama on a GPU node.

\section*{3 Evaluation Setup}

We evaluate the system in two complementary ways.

\textbf{Retrieval evaluation.} We created a small benchmark of 11 natural-language fitness queries with manually specified expectations. Each query defines relevant properties such as body part, equipment, difficulty level, and keywords. Retrieved items are marked relevant if they satisfy these constraints. We report Hit@1, Hit@3, Hit@5, Precision@5, and Mean Reciprocal Rank (MRR).

\textbf{End-to-end comparison.} We also compare baseline generation and RAG generation on five representative questions. This evaluation is qualitative rather than fully annotated. Its purpose is to show whether retrieval improves grounding, specificity, and alignment with the dataset.

\section*{4 Results}

Table~\ref{tab:main-results} summarizes the main quantitative results of the retrieval component.

\begin{table}[h]
\centering
\caption{Retrieval evaluation on 11 manually specified fitness queries. Higher is better for all metrics. The table is self-contained: Hit@$k$ measures whether at least one relevant exercise appears in the top $k$ retrieved items, Precision@5 measures the fraction of relevant items among the top five, and MRR measures how early the first relevant result appears.}
\label{tab:main-results}
\begin{tabular}{lr}
\toprule
Metric & Value \\
\midrule
Queries & 11 \\
Hit@1 & 0.636 \\
\textbf{Hit@3} & \textbf{0.909} \\
\textbf{Hit@5} & \textbf{0.909} \\
Precision@5 & 0.582 \\
MRR & 0.742 \\
\bottomrule
\end{tabular}
\end{table}

The results show that retrieval is already strong enough to support a useful RAG pipeline. In 10 out of 11 queries, at least one relevant exercise appears in the top five retrieved items. However, the lower Hit@1 score shows that the first retrieved result is still often suboptimal. This matters because the generator is sensitive to noisy or weak context.

Table~\ref{tab:qualitative} summarizes the qualitative comparison between baseline and RAG behavior.

\begin{table}[h]
\centering
\caption{Qualitative comparison between direct generation and RAG using the same Llama 3.2 1B model. The table highlights the main behavioral difference observed in the generated examples.}
\label{tab:qualitative}
\begin{tabular}{>{\raggedright\arraybackslash}p{3.1cm} >{\raggedright\arraybackslash}p{4.2cm} >{\raggedright\arraybackslash}p{4.2cm}}
\toprule
Aspect & Baseline & RAG \\
\midrule
Knowledge source & General model knowledge & Retrieved MegaGym entries + model \\

Exercise naming & Often generic or paraphrased & Usually matches dataset entries more closely \\

Grounding & No explicit grounding & Explicitly grounded in retrieved context \\

Failure mode & Unsupported but plausible suggestions & Wrong answers when retrieval is weak or noisy \\
\bottomrule
\end{tabular}
\end{table}

\section*{5 Analysis of Successes and Failures}

The clearest success case is the query \emph{``Can you recommend some triceps exercises that use a cable machine?''} In the baseline setting, the model gives plausible but generic cable-triceps suggestions. In the RAG setting, the answer is more tightly aligned with the dataset and mentions retrieved entries such as \emph{Cable Straight-Bar Push-Down} and \emph{Triceps Overhead Extension with Rope}. This is exactly the intended benefit of RAG: the same small model becomes more domain-grounded when given relevant context.

One representative failure case is \emph{``Which bodyweight leg exercises would you suggest for a beginner?''} Retrieval fails here because the returned items are only partially aligned with the intended constraints. Some retrieved exercises are plyometric, some have missing equipment information, and one result even targets abdominals rather than legs. As a consequence, the RAG answer inherits the retrieval error and becomes less useful than the baseline. This example is important because it shows that RAG does not automatically improve every answer; it improves answers only when retrieval is good enough.

We also observe partial failures caused by noisy metadata and duplicates. For example, some retrieved titles are near-duplicates, and some exercise descriptions are repeated or inconsistently labeled. This reduces Precision@5 and makes the prompt context less clean than it should be.

\section*{6 Discussion}

The main conclusion is that retrieval quality dominates final answer quality in this project. The generator is not the main bottleneck. Even with a small 1B model, the system can produce useful domain-specific answers when the retrieved context is relevant. Conversely, when retrieval is weak, the RAG answer degrades quickly.

This finding is practically useful for future work. The most promising improvements are not larger models, but better retrieval and cleaner data. In particular, the next steps should include:

\begin{itemize}
\item metadata-aware filtering by body part, equipment, and difficulty,
\item dataset cleaning and deduplication,
\item expansion of the evaluation benchmark beyond 11 retrieval queries,
\item more systematic human evaluation of final answers.
\end{itemize}

Our work is also consistent with the general motivation of RAG systems \cite{Lewis2020}: retrieval can inject task-specific knowledge into a fixed generator. In the fitness domain, related work has also explored exercise recommendation systems that combine structured information and NLP components \cite{Yang2023}. Our contribution is smaller in scale, but it demonstrates the same core principle in a reproducible course project setting.

\section*{7 Reproducibility and Final Take-Away}

The repository is structured so that another student can rerun the project with minimal setup. The dataset source is linked, indexing is scripted, evaluation is scripted, and both local and HPC execution paths are documented. This satisfies the course requirement that the project should be understandable and easy to rerun by reviewers.

\textbf{Take-away message.} With the same small 1B language model, RAG improves domain grounding by forcing answers to use retrieved MegaGym exercise entries. The main limitation is not the generator itself, but retrieval quality: when retrieval is strong, answers become more specific and trustworthy; when retrieval is weak, the whole system fails with it.

\bibliographystyle{unsrt}
\bibliography{report}

\end{document}